Response to reviewers

### 2024-11-30 Manuscript/ecography_reponse_letter.tex
final comment

%  @@@TODO: add stats for results (following other reviewers comments)

@@

\textcolor{blue}{Line 254-256. Apply the same changes mentioned in the comment below to lines 248-251.}
%  @@@TODO: add stats for results (following other reviewers comments)

@@

%  @@@TODO: add stats for results (following other reviewers comments)

@@

%  @@@TODO: Links back to main comment from reviewer 2: Go back to the previous version of table 1 that gives more description of the models
% One thing here is probably to justify more why we think that, for example, ``removing points from the side of a traitspace = filtering''.
%  We now changed the labels in the figure and the caption to highlight that we are indeed approximating such mechanisms, not simulating them per se as the reviewers pointed out.

>>>>

We have now added all the statistics in the supplementary materials table 2 to 5 (for the model results) and 6 to 18 (for the metrics differences to the nulls results).

@@

\textcolor{blue}{Line 254-256. Apply the same changes mentioned in the comment below to lines 248-251.}

We now refer to the added statistics in the supplementary materials table 2 to 5 (for the model results) and 6 to 18 (for the metrics differences to the nulls results).

@@

@@

We have now improved both the mechanisms approximation description (see comment below) as well as added more justification on why we believe our algorithms approximate the focal mechanisms (see response to the next comment). Furthermore, we have also added some general predictions as to which type of metric would capture which type of changes in the trait space (Table 1). We hope that these three points improve the clarity of our objectives and justifications.
We now changed the labels in the figure and the caption to highlight that we are indeed approximating such mechanisms, not simulating them \textit{per se} as the reviewers pointed out.

### 2024-11-30 Manuscript/main_ecography_R1.tex
final comment

First, we simulated a single random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e. a pure birth speciation model) until reaching 200 observations in \texttt{R} \citep{rcore} using \texttt{treats} \citep{guillerme2024treats}.
This trait was simulated either as a 2, 4 or 8 dimensional trait, in other words a Brownian Motion simulated in a 2, 4 or 8 dimensions.
Note here that we refer to traits as any measurable aspect of an organism (\textit{sensu} \cite{mcgill2006rebuilding} - but see \cite{dawson2021traits} for nuances in what researchers perceive).
These aspects are often expressed in one dimension but can in fact be described in any number of dimensions.
For example, for trait defined as ``leaf insertion angle'', this can measured in one dimension (an angle in degrees) or three dimensions (the same angle expressed as the trigonometric relation of three sides of a triangle).
These observations on which the traits are measured can represent any biological focal entities, e.g. tips in a phylogenetic tree, individuals, species, populations, OTUs, etc. (see Glossary table in supplementary materials).
This resulted in a neutral null model of trait evolution with no effect of competition, extinction, selection or other processes (\textit{sensu} \citealt{bausman2018modeling}).
We refer to this as the ``non-stressed trait space''.

@@

The ability of different metrics to capture the different patterns (and thus approximate the mechanisms) was highly variable.
It ranged from metrics capturing no clear pattern (e.g. the Functional evenness based on the dendrogram method for competition) to metrics clearly capturing one specific pattern (e.g. Divergence based on probability density for the equalising mechanism for a decrease in space occupancy).
Although most metrics captured a decrease in metric score relative to the amount of removed data, some metrics resulted in a score increase (e.g. the Functional evenness based on a hypervolume for the equalising mechanism) or non-linear responses (e.g. Alpha diversity based on a dendrogram for the equalising epresent distributions of metric scores).
Some cases resulted in an increase in metric score relative to random removals (positive values in Figure \ref{Fig:simulation_results}).
Furthermore, although we predicted some more common changes in aspects of functional diversity for some specific mechanisms, our simulations show that most metrics under most aspects (richness, divergence or regularity) capture changes of trait space occupancy under any mechanism (see caveats section in the discussion).

@@

For the empirical data, we see again a range of different possible interpretations depending on the metric describing the pattern (Figure \ref{Fig:empirical_results}, Supplementary table 5).
For example, all richness metrics (green; Figure \ref{Fig:empirical_results}) pick up a similar relative increase (i.e., relative to the null stressor) in richness for both the historic and extant datasets.
However, the results are occasionally contrasting between the historical and extant datasets, with a respective decrease and increase with some divergence metrics (e.g. functional dispersion based on dendrograms or hypervolumes).
Finally, some metrics pick up a decrease for both datasets (e.g. functional evenness based on hypervolumes - although the $95$\% confidence intervals overlapped with $0$).
Note that this apparent global increase in most metric values may appear counter intuitive given that $46$\% and $69$\% of species went extinct over the two time periods.
This is due to the scaling of the metrics compared to random equivalent extinctions.
That is, Figure \ref{Fig:empirical_results} is not displaying absolute changes in trait space occupancy but rather changes in trait space occupancy relative to a crude neutral model where all species are equally likely to go extinct.

>>>>

First, we simulated a single random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e. a pure birth speciation model) until reaching 200 observations in \texttt{R} \citep{rcore} using \texttt{treats} \citep{guillerme2024treats}.
This trait was simulated either as a 2, 4 or 8 dimensional trait, in other words a Brownian Motion simulated in a 2, 4 or 8 dimensions.
Note here that we refer to traits as any measurable aspect of an organism (\textit{sensu} \cite{mcgill2006rebuilding} - but see \cite{dawson2021traits} for nuances in what researchers perceive).
These aspects are often expressed in one dimension but can in fact be described in any number of dimensions.
For example, for trait defined as ``leaf insertion angle'', this can measured in one dimension (an angle in degrees) or three dimensions (the same angle expressed as the trigonometric relation of three sides of a triangle).
These observations on which the traits are measured can represent any biological focal entities, e.g. tips in a phylogenetic tree, individuals, species, populations, OTUs, etc. (see Glossary in the supplementary materials, table 1).
This resulted in a neutral null model of trait evolution with no effect of competition, extinction, selection or other processes (\textit{sensu} \citealt{bausman2018modeling}).
We refer to this as the ``non-stressed trait space''.

@@

The ability of different metrics to capture the different patterns (and thus approximate the mechanisms) was highly variable.
It ranged from metrics capturing no clear pattern (e.g. the Functional evenness based on the dendrogram method for competition) to metrics clearly capturing one specific pattern (e.g. Divergence based on probability density for the equalising mechanism for a decrease in space occupancy).
Although most metrics captured a decrease in metric score relative to the amount of removed data, some metrics resulted in a score increase (e.g. the Functional evenness based on a hypervolume for the equalising mechanism) or non-linear responses (e.g. Alpha diversity based on a dendrogram for the equalising epresent distributions of metric scores; see supplementary tables 2 to 5 and 6 to 17).
Some cases resulted in an increase in metric score relative to random removals (positive values in Figure \ref{Fig:simulation_results}).
Furthermore, although we predicted some more common changes in aspects of functional diversity for some specific mechanisms, our simulations show that most metrics under most aspects (richness, divergence or regularity) capture changes of trait space occupancy under any mechanism (see caveats section in the discussion).

@@

For the empirical data, we see again a range of different possible interpretations depending on the metric describing the pattern (Figure \ref{Fig:empirical_results}, Supplementary table 5 and 18).
For example, all richness metrics (green; Figure \ref{Fig:empirical_results}) pick up a similar relative increase (i.e., relative to the null stressor) in richness for both the historic and extant datasets.
However, the results are occasionally contrasting between the historical and extant datasets, with a respective decrease and increase with some divergence metrics (e.g. functional dispersion based on dendrograms or hypervolumes).
Finally, some metrics pick up a decrease for both datasets (e.g. functional evenness based on hypervolumes - although the $95$\% confidence intervals overlapped with $0$).
Note that this apparent global increase in most metric values may appear counter intuitive given that $46$\% and $69$\% of species went extinct over the two time periods.
This is due to the scaling of the metrics compared to random equivalent extinctions.
That is, Figure \ref{Fig:empirical_results} is not displaying absolute changes in trait space occupancy but rather changes in trait space occupancy relative to a crude neutral model where all species are equally likely to go extinct.
